%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------


\documentclass[preprint,9pt,twocolumn,nocopyrightspace,authoryear]{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{common}
\usepackage{minted}
\include{bibliography.bib}
\usepackage{chngcntr}

\newcounter{ex}
\newenvironment{exmp}{
\refstepcounter{ex}%
\par\vspace{.8em}\hfill\framebox{\small Example~\theex}\nopagebreak
\\[-2.4em]\nopagebreak}
{\relax}


\begin{document}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{20yy} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
\doi{nnnnnnn.nnnnnnn}

\newcommand{\clj}[1]{\mintinline{clojure}{#1}}
\newcommand{\java}[1]{\mintinline{java}{#1}}
\newcommand{\rkt}[1]{\mintinline{racket}{#1}}

% Uncomment one of the following two, if you are not going for the 
% traditional copyright transfer agreement.

%\exclusivelicense                % ACM gets exclusive license to publish, 
                                  % you retain copyright

%\permissiontopublish             % ACM gets nonexclusive license to publish
                                  % (paid open-access papers, 
                                  % short abstracts)

%\titlebanner{banner above paper title}        % These are ignored unless
%\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Practical Optional Types for Clojure}

\authorinfo{Ambrose Bonnaire-Sergeant}
           {Indiana University}
           {abonnair@indiana.edu}
\authorinfo{Rowan Davies}
           {University of Western Australia}
           {rowan.davies@uwa.edu.au}
\authorinfo{Sam Tobin-Hochstadt}
           {Indiana University}
           {samth@indiana.edu}

\maketitle

\begin{abstract}
Typed Clojure is an optional type system for Clojure, a dynamic
language in the Lisp family that targets the JVM. Typed Clojure's type
system build on the design of Typed Racket, repurposing in particular
\emph{occurrence typing}, its approach to statically reasoning about
predicate tests. However, in adapting the type system to Clojure,
changes and extensions are required to accommodate both additional
language features and  idioms used by Clojure programmers.

In this paper, we describe Typed Clojure and present these type system
extensions, focusing on three features widely used in Clojure. First, Java
interoperability is central to Clojure's mission but introduces
challenges such as ubiquitous \texttt{null}; Typed Clojure handles
Java interoperability while ensuring the absence of null-pointer
exceptions in typed programs. Second, Clojure programmers
idiomatically use immutable dictionaries for data structures; Typed
Clojure handles this in the type system with multiple forms of
heterogeneous dictionary types. Third, multimethods provide extensible
operations, and their Clojure semantics turns out to have a surprising
synergy with the underlying occurrence typing framework.

We provide a formal model of the Typed Clojure type system
incorporating these and other features, with a proof of
soundness. Additionally, Typed Clojure is now in use by numerous
corporations and developers working with Clojure, and we report on 
experience with the system and its lessons for the future.


%% Clojure is a dynamically typed language hosted on the Java 
%% Virtual Machine.
%% Typed Racket is a valuable starting point for
%% a gradual type system that targets Clojure.
%% Building a similar type system for a new language gives the
%% designer some flexibility to repurpose and extend features.
%% This paper gives an overview of Typed Clojure, concentrating
%% on the extensions and differences from Typed Racket. We also
%% show where Typed Racket's features were particularly useful
%% for type checking non-trivial Clojure idioms.
\end{abstract}

%% \category{CR-number}{subcategory}{third-level}

%% \keywords
%% optional types, Clojure

\input{intro}

\input{overview}

\input{formal-model}

\input{metatheory}

\input{experience}

\input{related-work}

\input{conclusion}

\input{acknowledgements}

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

\bibliography{bibliography}

\clearpage

\input{appendix}

\end{document}

%                       Revision History
%                       -------- -------
%  Date         Person  Ver.    Change
%  ----         ------  ----    ------

%  2013.06.29   TU      0.1--4  comments on permission/copyright notices

